---
title: "community assembly analysis for the CTS dataset \n(the rare and dominant biosphere)"
author: "Xiu Jia (xibeihenai@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

The composted tannery sludge (CTS) dataset are from Prof. Ademir Araujo <asfaruaj@yahoo.com.br> from Federal University of Piauí. 
Sequence data was analyzed by Dr. Lucas William Mendes <lucaswmendes@gmail.com> from University of Sao Paulo.

***

##### Treatment by composted tannery sludge (CTS):
* T1 - 0 ton CTS/ha
* T2 - 2.5 ton CTS/ha
* T3 - 5 ton CTS/ha
* T4 - 10 ton CTS/ha
* T5 - 20 ton CTS/ha

##### Sampling time points:
* 0 - 1 days after CTS application
* 45 - 45 days after CTS application
* 75 - 75 days after CTS application
* 150 - 150 days after CTS application
* 180 - 180 days after CTS application


### initiate libraries
```{r setup, message = FALSE, warning = FALSE}

rm(list=ls())

library(tidyverse)
library(corrplot)
library(ggforce)
library(RColorBrewer)


mytheme <- theme_bw()+
  theme(panel.spacing = unit(0, "lines"),
        strip.background = element_blank(),
        strip.placement = "outside",
        legend.box.background = element_rect(),
        legend.position = "right", 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

# make symmetric matrix from lower diagonal [cite: https://stackoverflow.com/questions/33026183/r-make-symmetric-matrix-from-lower-diagonal]
makeSymm <- function(m) {
  m[upper.tri(m)] <- t(m)[upper.tri(m)]
  return(m)
}

```


### overall assembly processes structuring the rare and dominant biospheres
```{r}

# names of two datasets 
data.set.names <- c("dominant", "rare") 

# make a matrix to store the number of pairwise samples of each assembly process
df <- matrix(NA, nrow = 2, ncol = 5)
colnames(df) <- c("Variable.selection", "Homogeneous.selection", 
                   "Dispersal.limitation", "Homogenizing.dispersal", "Undominated.processes")
rownames(df) <- data.set.names
df

# a loop to calculte for the assembly processes of each dataset 
for (data.set.name in data.set.names) {
  
  cat("\nFor dataset:", data.set.name, "\n")
  
  
  ### load datasets & visualize the matrices
  
  
  print("beta-NTI")
  
  nti_raw <- read.csv(paste('betaNTI_weighted_', data.set.name, ".csv", sep = ""), header = 1, row.names = 1, check.names = FALSE)
  
  nti_raw <- as.matrix(nti_raw)
  
  nti_sym <- makeSymm(nti_raw)
  
  nti_sym[is.na(nti_sym)] <- 0
  
  nti_sym <- as.matrix(nti_sym)
  
  pdf(file = paste("beta_nit_matrix_", data.set.name, ".pdf", sep = ""))
  
  corrplot(nti_sym, is.corr = FALSE, type = "lower", diag = F, 
           order = "original", 
           col=colorRampPalette(c("blue4", "white", "firebrick1"))(100), 
           tl.col = "black", tl.srt = 45, tl.cex = 0.5,
           title = paste("beta-NTI matrix of the",  data.set.name, "biosphere"), 
           mar = c(0,0,1,0), number.cex = 0.5, number.digits = 2)
  
  dev.off()
  
  
  
  print("RC-bray")
  
  rc_raw <- read.csv(paste('RC-bray_', data.set.name, ".csv", sep = ""), header = 1, row.names = 1, check.names = FALSE)
  
  rc_raw <- as.matrix(rc_raw)
  
  rc_sym <- makeSymm(rc_raw)
  
  rc_sym[is.na(rc_sym)] <- 0
  
  rc_sym <- as.matrix(rc_sym)
  
  
  pdf(file = paste("RC-bray_matrix_", data.set.name, ".pdf", sep = ""))
  
  corrplot(rc_sym, is.corr = FALSE, type = "lower", diag = F, 
           order = "original", 
           col=colorRampPalette(c("blue4", "white", "firebrick1"))(100), 
           tl.col = "black", tl.srt = 45, tl.cex = 0.5,  cl.cex = 0.5,
           title = paste("RC-bray matrix of the",  data.set.name, "biosphere"), 
           mar = c(0,0,1,0))
  
  dev.off()
  
  
  
  
  ### expand weighted beta NTI
  
  # extract pairwise beta-NTI from a n*n lower trianglar matrix
  nti <- data.frame(as.table(nti_raw))[lower.tri(nti_raw, diag = FALSE), ]
  
  cat("NO. of NTI pairs should be", (nrow(nti_raw)*nrow(nti_raw)-nrow(nti_raw))/2 , 
      "\nIs Observed = Predict?", (nrow(nti_raw)*nrow(nti_raw)-nrow(nti_raw))/2 == length(nti$Freq), 
      "\nthe average of beta-NTI is:", round(mean(na.omit(nti$Freq)),2))
  
  row.names(nti) <- paste(nti$Var1, nti$Var2, sep = "_")
  
  # head(nti)
  # str(nti)
  
  
  # extract pairwise RC-bray from a n*n lower trianglar matrix
  rc <- data.frame(as.table(rc_raw))[lower.tri(rc_raw, diag = FALSE), ]
  
  cat("NO. of RC-bray pairs should be", (nrow(rc_raw)*nrow(rc_raw)-nrow(rc_raw))/2 , 
      "\nIs Observed = Predict?", (nrow(rc_raw)*nrow(rc_raw)-nrow(rc_raw))/2 == length(rc$Freq), 
      "\nthe average of RC-bray is:", round(mean(na.omit(rc$Freq)),2))
  
  row.names(rc) <- paste(rc$Var1, rc$Var2, sep = "_")
  group <- data.frame(row.names = rownames(rc), t(as.data.frame(strsplit(as.character(row.names(rc)), "_"))))
  
  # head(rc)
  # str(rc)
  
  
  # Combine beta-NTI and RC-bray
  nti.rc <- merge(nti, rc, by = "row.names", all = TRUE)  
  
  nti.rc <- data.frame(nti = nti.rc$Freq.x, rc = nti.rc$Freq.y, row.names = nti.rc$Row.names)
  
  # Invalid the value of RC-bray in which the beta-NTI larger than +2 or less than -2
  for (i in 1:nrow(nti.rc)) {
    if (nti.rc[i, 1] > 2 | nti.rc[i, 1] < -2) {
      nti.rc[i, 2] <- NA
    }
  }
  
  # head(nti.rc)
  str(nti.rc)
  
  
  # Quantify each assembly process 
  if (data.set.name == "dominant") { 
    i = 1 } else if (data.set.name == "rare") {
      i = 2
    } 
  
  # Variable selection
  Variable.selection <- nti.rc$nti > 2
  cat('variable  selection:', table(Variable.selection)['TRUE'], 'within', nrow(nti.rc), 'pairwise samples\n')
  df[i, 1] <- length(Variable.selection[Variable.selection == TRUE])
  
  # Homogeneous selection
  Homogeneous.selection <- nti.rc$nti < -2
  cat('homogenous selection:', table(Homogeneous.selection)['TRUE'], 'within', nrow(nti.rc), 'pairwise samples\n')
  df[i, 2] <- table(Homogeneous.selection)['TRUE']#length(c[Homogeneous.selection == TRUE])
  
  # Dispersal limitation
  Dispersal.limitation <- na.omit(nti.rc$rc) > 0.95
  cat('dispersal limitation:', table(Dispersal.limitation)['TRUE'], 'within', nrow(nti.rc), 'pairwise samples\n')
  df[i, 3] <- length(Dispersal.limitation[Dispersal.limitation == TRUE])
  
  # Homogenizing dispersal
  Homogenizing.dispersal <- na.omit(nti.rc$rc) < -0.95
  cat('homogenizing dispersal:', table(Homogenizing.dispersal)['TRUE'], 'within', nrow(nti.rc), 'pairwise samples\n')
  df[i, 4] <- length(Homogenizing.dispersal[Homogenizing.dispersal == TRUE])
  
  # Undominated processes
  Undominated.processes <- na.omit(nti.rc$rc) <= 0.95 & na.omit(nti.rc$rc) >= -0.95
  cat('Undominated processes:', table(Undominated.processes)['TRUE'], 'within', nrow(nti.rc), 'pairwise samples\n')
  df[i, 5] <- length(Undominated.processes[Undominated.processes == TRUE])
  
}



df <- df %>% 
  as.data.frame %>% 
  mutate_if(is.character, as.numeric) 

df

df1 <- apply(df, 1, function(x) x * 100 / sum(x)) %>% 
  as.data.frame %>% 
  rownames_to_column(var = "Assembly_processes") %>% 
  pivot_longer(-Assembly_processes, names_to = "Datasets", values_to = "Value") %>% 
  mutate(Assembly_processes = factor(Assembly_processes, levels = c("Variable.selection", "Homogeneous.selection", "Dispersal.limitation", "Homogenizing.dispersal", "Undominated.processes"),  labels = c("Variable selection", "Homogeneous selection", "Dispersal limitation",  "Homogenizing dispersal", "Undominated processes"))) %>% 
  mutate(Datasets = factor(Datasets, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere")))

head(df1)

# first way to generate a pie plot
# clockwise
# calculate the start and end angles for each pie
dat_pies <- left_join(df1,
                      df1 %>% 
                        group_by(Datasets) %>%
                        summarize(value_total = sum(Value))) %>%
  group_by(Datasets) %>%
  mutate(end_angle = 2*pi*cumsum(Value)/value_total,      # ending angle for each pie slice
         start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
         mid_angle = 0.5*(start_angle + end_angle))   # middle of each pie slice, for the text label

rpie = 1 # pie radius
rlabel = 0.6 * rpie # radius of the labels; a number slightly larger than 0.5 seems to work better, 0.5 would place it exactly in the middle as the question asks for.

# draw the pies
(pie <- ggplot(dat_pies) + 
  geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = rpie, start = start_angle, end = end_angle, fill = Assembly_processes)) +
  geom_text(aes(x = rlabel*sin(mid_angle), y = rlabel*cos(mid_angle), label = paste(round(Value,2), "%")), 
            hjust = 0.5, vjust = 0.5, size=4) +
  coord_fixed() +
  scale_x_continuous(limits = c(-1, 1), name = "", breaks = NULL, labels = NULL) +
  scale_y_continuous(limits = c(-1, 1), name = "", breaks = NULL, labels = NULL) +
  facet_wrap(. ~ Datasets, ncol = 2)+
  scale_fill_manual(values = c("#00A087B2","#FFDB6D", "#DC0000B2", "#4DBBD5B2", "#7570B3"), name = "Assembly processes") +
  theme_minimal())

# ggsave("Assembly_processes_pie_plot_raw.pdf", width = 9, height = 3.5, units = "cm", pie, scale = 2)

```


### assembly processes structuring the temporal variation
```{r}


# names of two datasets 
data.set.names <- c("dominant", "rare") 

# make a list to store the assembly process results of each dataset
datalist <- list()

# a loop to calculte for the assembly processes of each dataset 
for (data.set.name in data.set.names) {
  
  cat("For dataset:", data.set.name, "\n")
  
  
  # load datasets
  
  nti_raw <- read.csv(paste('betaNTI_weighted_', data.set.name, ".csv", sep = ""), header = 1, row.names = 1, check.names = FALSE)
  
  rc_raw <- read.csv(paste('RC-bray_', data.set.name, ".csv", sep = ""), header = 1, row.names = 1, check.names = FALSE)
  
  
  nti_raw <- as.matrix(nti_raw)
  
  rc_raw <- as.matrix(rc_raw)
  
  
  ### expand weighted beta NTI
  
  # extract pairwise beta-NTI from a n*n lower trianglar matrix
  nti <- data.frame(as.table(nti_raw))[lower.tri(nti_raw, diag = FALSE), ]
  
  cat("NO. of NTI pairs should be", (nrow(nti_raw)*nrow(nti_raw)-nrow(nti_raw))/2 , 
      "\nIs Observed = Predict?", (nrow(nti_raw)*nrow(nti_raw)-nrow(nti_raw))/2 == length(nti$Freq), 
      "\nthe average of beta-NTI is:", round(mean(na.omit(nti$Freq)),2))
  
  row.names(nti) <- paste(nti$Var1, nti$Var2, sep = "_")
  group <- data.frame(row.names = rownames(nti), t(as.data.frame(strsplit(as.character(row.names(nti)), "_"))))
  group$X2 <- gsub('.{1}$', '', group$X2)
  group$X4 <- gsub('.{1}$', '', group$X4)
  
  # add group information
  nti <- data.frame(row.names = rownames(nti), 
                    X1 = group$X1,
                    X2 = group$X2,
                    X3 = group$X3, 
                    X4 = group$X4, 
                    Freq = nti$Freq)
  
  # keep pairs with same treatment
  nti <- nti[which(nti$X1==nti$X3), ]
  
  # ignore pairs samping at the same day
  nti <- nti[which(nti$X2!=nti$X4), ]
  
  head(nti)
  str(nti)
  
  
  # extract pairwise RC-bray from a n*n lower trianglar matrix
  rc <- data.frame(as.table(rc_raw))[lower.tri(rc_raw, diag = FALSE), ]
  
  cat("NO. of rc pairs should be", (nrow(rc_raw)*nrow(rc_raw)-nrow(rc_raw))/2 , 
      "\nIs Observed = Predict?", (nrow(rc_raw)*nrow(rc_raw)-nrow(rc_raw))/2 == length(rc$Freq), 
      "\nthe average of RC-bray is:", round(mean(na.omit(rc$Freq)),2))
  
  row.names(rc) <- paste(rc$Var1, rc$Var2, sep = "_")
  group <- data.frame(row.names = rownames(rc), t(as.data.frame(strsplit(as.character(row.names(rc)), "_"))))
  group$X2 <- gsub('.{1}$', '', group$X2)
  group$X4 <- gsub('.{1}$', '', group$X4)
  
  # add group information
  rc <- data.frame(row.names = rownames(rc), 
                   X1 = group$X1,
                   X2 = group$X2,
                   X3 = group$X3, 
                   X4 = group$X4, 
                   Freq = rc$Freq)
  
  # keep pairs with same treatment
  rc <- rc[which(rc$X1==rc$X3), ]
  
  # ignore pairs samping at the same day
  rc <- rc[which(rc$X2!=rc$X4), ]
  
  head(rc)
  str(rc)
  
  
  # Combine beta-NTI and RC-bray
  nti.rc <- merge(nti, rc, by = "row.names", all = TRUE)  
  
  nti.rc <- data.frame(Treat = nti.rc$X1.x, nti = nti.rc$Freq.x, rc = nti.rc$Freq.y, row.names = nti.rc$Row.names)
  
  # Invalid the value of RC-bray in which the beta-NTI larger than +2 or less than -2
  for (i in 1:nrow(nti.rc)) {
    if (nti.rc[i,2] > 2 | nti.rc[i,2] < -2) {
      nti.rc[i, 3] <- NA
    }
  }
  
  nti.rc <- nti.rc %>%
    mutate(Treat = factor(Treat, 
                          levels = c("T1", "T2", "T3", "T4", "T5"), 
                          labels = c("0 ton", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha")))
  
  head(nti.rc)
  str(nti.rc)
  
  
  # calculate the relative influence of different assembly processes for each treatment
  
  # make a matrix to store the number of pairwise samples of each assembly process
  df <- matrix(NA, nrow = 5, ncol = 6)
  
  df[, 1] <- c("0 ton", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha")
  
  colnames(df) <- c("Treat", "Variable.selection", "Homogeneous.selection", 
                    "Dispersal.limitation", "Homogenizing.dispersal", "Undominated.processes")
  
  
  
  for (Treat in levels(nti.rc$Treat) ) {
    
    cat("for treatment of ", Treat, ":\n")
    
    # Variable selection ---
    Variable.selection <- nti.rc[nti.rc$Treat == Treat, ]$nti > 2
    
    cat("Number of variable  selection :",  table(Variable.selection)["TRUE"], 
        "within", nrow(nti.rc[nti.rc$Treat == Treat, ]), "pairwise samples\n")
    
    df[df[,1] == Treat, 2] <- length(Variable.selection[Variable.selection == TRUE])
    
    # Homogeneous selection ---
    Homogeneous.selection <- nti.rc[nti.rc$Treat == Treat, ]$nti < -2
    
    cat("Number of Homogeneous selection is:", table(Homogeneous.selection)["TRUE"], 
        "within", nrow(nti.rc[nti.rc$Treat == Treat, ]), "pairwise samples\n")
    
    df[df[,1] == Treat, 3] <-  length(Homogeneous.selection[Homogeneous.selection == TRUE]) # table(Homogeneous.selection)["TRUE"] #
    
    # Dispersal limitation ---
    Dispersal.limitation <- na.omit(nti.rc[nti.rc$Treat == Treat, ]$rc) > 0.95
    
    cat("Number of dispersal limitation is:", table(Dispersal.limitation)["TRUE"], 
        "within", nrow(nti.rc[nti.rc$Treat == Treat, ]), "pairwise samples\n")
    
    df[df[,1] == Treat, 4] <- length(Dispersal.limitation[Dispersal.limitation == TRUE])
    
    # Homogenizing dispersal ---
    Homogenizing.dispersal <- na.omit(nti.rc[nti.rc$Treat == Treat, ]$rc) < -0.95
    
    cat("Number of homogenizing dispersal is:", table(Homogenizing.dispersal)["TRUE"], 
        "within", nrow(nti.rc[nti.rc$Treat == Treat, ]), "pairwise samples\n")
    
    df[df[,1] == Treat, 5] <- length(Homogenizing.dispersal[Homogenizing.dispersal == TRUE])
    
    # Undominated processes ---
    Undominated.processes <- na.omit(nti.rc[nti.rc$Treat == Treat, ]$rc) <= 0.95 & 
      na.omit(nti.rc[nti.rc$Treat == Treat, ]$rc) >= -0.95
    
    cat("Number of Undominated processes is:", table(Undominated.processes)["TRUE"], 
        "within", nrow(nti.rc[nti.rc$Treat == Treat, ]), "pairwise samples\n")
    
    df[df[,1] == Treat, 6] <- length(Undominated.processes[Undominated.processes == TRUE])
  }
  
  datalist[[data.set.name]] <- df
}

str(datalist)

df <- do.call(rbind.data.frame, datalist)

df

df$Datasets <- factor(gsub(".{2}$", "", row.names(df)))

df <- df %>% 
  as.data.frame()

df[, -c(1,7)] <- sapply(df[, -c(1, 7)], as.numeric)

df1 <- df %>% 
  # as.data.frame %>%
  mutate(across(-c(1, 7))*100/rowSums(across(-c(1, 7)))) %>%
  pivot_longer(!c(Treat, Datasets), names_to = "Assembly_processes", values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("0 ton", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Assembly_processes = factor(Assembly_processes, levels = c("Variable.selection", "Homogeneous.selection", "Dispersal.limitation", "Homogenizing.dispersal", "Undominated.processes"),  labels = c("Variable selection", "Homogeneous selection", "Dispersal limitation",  "Homogenizing dispersal", "Undominated processes"))) %>% 
  mutate(Datasets = factor(Datasets, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere")))

head(df1)

# stacked-bar plot of the assembly processes structuring the temporal variation of the dominant/rare biosphere
(p <- ggplot(df1, aes(x = Treat, y = value, fill = Assembly_processes)) + 
    geom_bar(stat = "identity", width = 0.8, colour = "black") +
    scale_y_continuous(expand = c(0, 0), limits = c(0,105))+
    facet_grid(.~ Datasets) +
    scale_fill_manual(values = c("#00A087B2","#FFDB6D", "#DC0000B2", "#4DBBD5B2", "#7570B3"), name = "Assembly processes") +
    labs(x = "", y = "Relative Influence (%)", title = "") +
    mytheme +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5)))

# ggsave("Assembly_processes_rare_dominant_biospheres.pdf", width = 12, height = 7, units = "cm", p, scale = 1.5)

```


### Session Info
```{r}
sessionInfo()
```