---
title: "Rare biosphere analysis for the CTS dataset"
author: "Xiu Jia (xibeihenai@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

The composted tannery sludge (CTS) dataset are from Prof. Ademir Araujo <asfaruaj@yahoo.com.br> from Federal University of Piauí. 
Sequence data was analyzed by Dr. Lucas William Mendes <lucaswmendes@gmail.com> from University of Sao Paulo.

***

##### Treatment by composted tannery sludge (CTS):
* T1 - 0 ton CTS/ha
* T2 - 2.5 ton CTS/ha
* T3 - 5 ton CTS/ha
* T4 - 10 ton CTS/ha
* T5 - 20 ton CTS/ha

##### Sampling time points:
* 0 - 1 days after CTS application
* 45 - 45 days after CTS application
* 75 - 75 days after CTS application
* 150 - 150 days after CTS application
* 180 - 180 days after CTS application


### initiate libraries
```{r setup, message=FALSE, warning=FALSE}

rm(list=ls())

library(tidyverse)
library(vegan)
library(picante)
library(reshape2)
library(scales) # to access break formatting functions
library(RColorBrewer)
# display.brewer.all()

```


### load OTU table
```{r}

# feature table
com <- read.csv("out_table_rarefied.csv", header = 1, row.names = 1, sep = ",") 
com <- t(com)
str(com)

knitr::kable(
  com[1:5, 1:2],
  caption = "how the OTU table looks"
)

cat("the OTU/feature table has been rarefied to", rarefactiondepth <- mean(rowSums(com)))

```
The number of samples is: `r ncol(com)`; the number of species/ASVs is: `r nrow(com)`; the range of sequence number among samples is: `r range(colSums(com))`.


### generate matrix of rank abundance curves
```{r}

# Calculate the lowest relative abundance (%) of the maximum ESV occurance in all samples
row_maximum <-  apply(com, 1, max)

warning("For this dataset, rarity cutoff should not more than ", round(min(row_maximum)/rarefactiondepth*100, 2), "% of the total abundance per sample!")


# source function
source("rad.matrix.R")

# generate a rank abundance matrix
rad <- rad.matrix(t(com))
row.names(rad) <- rad[ ,1]
rad <- rad[, -1]
df <- melt(t(rad))
df <- df[order(df$Var1),]

df <- df %>% separate(Var1, into = c("Treatment", "group"), sep = "_", remove = FALSE, fill = "right", convert = TRUE) %>% 
  separate(group, into = c("Day", "replicate"), sep = 1, fill = "right")

head(df)

```

### find sample-specific rarity cut-offs
```{r}
# using method calculating H-index to get the rarity cutoff per sample
Chao <- as.data.frame(t(estimateR(com)))
Chao$slope <- Chao$S.obs/Chao$S.chao1
head(Chao)

# built empty matrix
cutoffs <- matrix(NA, nrow(com), 3)
row.names(cutoffs) <- row.names(com)
cutoffs[,2] <- Chao$slope
cutoffs[,3] <- Chao$S.obs
colnames(cutoffs) <- c("Rarity.cutoffs", "Slopes", "S.obs")

# find rarity cutoffs 
for (j in 1:nrow(com)) {
  com_j <- sort(as.numeric(com[j, ]), decreasing = TRUE)
  com_j <- com_j[com_j!=0]
  slope <- cutoffs[j, 2]
  for (i in 1:length(com_j)){
    if (com_j[i] >= i*slope){
      H=i
    }
  }
  cutoffs[j, 1] <- H
}

cutoffs <- as.data.frame(cutoffs)
cutoffs$Var1 <- factor(row.names(cutoffs))

head(cutoffs)

```
The average rarity cutoff of sample-specific rarity cut-offs is `r mean(cutoffs$Rarity.cutoffs)*100/rarefactiondepth`%.


### rank abundance curves with sample specific cutoffs
```{r}

# only show part of the rank abundance curves
(p <- ggplot(df, aes(x = Var2, y = value, color = Treatment)) +
    geom_line(size = 0.4, alpha = 0.6) +
    facet_wrap(. ~ Var1, nrow = 9) +
    geom_text(data = cutoffs, mapping = aes(x = -Inf, y = -Inf, label = Rarity.cutoffs),
              size=3, color = "#006699", hjust = -0.5, vjust = -2) +
    scale_y_continuous(limits = c(0, 100)) +
    scale_x_continuous(limits = c(0, 100)) +
    geom_abline(data = cutoffs, aes(slope = Slopes, intercept = 0*Slopes, color = "#0099CC")) +
    labs(x = "Rank", y = "Reads", title= "") +
    theme_bw() +
    theme(text = element_text(size = 8),
          strip.background = element_blank(),
          strip.placement = "outside",
          strip.text = element_text(face = "bold"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()))

# ggsave("Rank_abundance_curve_zoomin_2.pdf", width = 20, height = 20, units = "cm", p, scale = 1.5)


# show the whole rank abundance curves
(p <- ggplot(df, aes(x = Var2, y = value, color = Treatment)) +
    geom_line(size = 0.4, alpha = 0.6) +
    facet_wrap(. ~ Var1, nrow = 9) +
    geom_text(data = cutoffs, mapping = aes(x = -Inf, y = -Inf, label = Rarity.cutoffs),
              size=3, color = "#006699", hjust = -0.3, vjust = -4) +
    geom_abline(data = cutoffs, aes(slope = Slopes, intercept = 0*Slopes, color = "#0099CC")) +
    labs(x = "Rank", y = "Reads", title= "") +
    theme_bw() +
    theme(text = element_text(size = 8),
          strip.background = element_blank(),
          strip.placement = "outside",
          strip.text = element_text(face = "bold"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()))

# ggsave("Rank_abundance_curve_ssrc.pdf", width = 20, height = 20, units = "cm", p, scale = 1.5)

```


### rank abundance curves with all rarity cutoffs
Three commonly used cutoffs are: 0.01% - `r rarefactiondepth*0.0001` reads, 0.1% - `r rarefactiondepth*0.001` reads and 1% - `r rarefactiondepth*0.01` reads.
```{r}

c(brewer.pal(9, "Blues")[c(3, 5, 7)])

# y log-scale
(p <- ggplot(df, aes(x = Var2, y = value, group = Var1)) +
    geom_line(size = 0.4, alpha = 0.6, color = "gray") +
    scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                  labels = trans_format("log10", math_format(10^.x))) +
    geom_hline(yintercept = c(1.45), linetype = "dashed", size = 0.4, color = "#C6DBEF") + 
    geom_hline(yintercept = c(14.5), linetype = "dashed", size = 0.4, color = "#6BAED6") + 
    geom_hline(yintercept = c(1450), linetype = "dashed", size = 0.4, color = "#2171B5") + 
    geom_hline(yintercept = c(mean(cutoffs$Rarity.cutoffs)), linetype = "dashed", size = 0.4, color = "#FFCC99") + 
    annotate("text", x = 2800, y = 2, color = "#C6DBEF", label = "Rarity cutoff < 0.01 %", size = 6/.pt) + 
    annotate("text", x = 2800, y = 20, color = "#6BAED6", label = "Rarity cutoff < 0.1 %", size = 6/.pt) +
    annotate("text", x = 2800, y = 1800, color = "#2171B5", label = "Rarity cutoff < 1.0 %", size = 6/.pt) +
    annotate("text", x = 2800, y = 60, color = "#FFCC99", label = "Sample-specific rarity cutoffs ~ 0.3 %", size = 6/.pt) + 
    labs(x = "Rank", y= "Reads", title= "") +
    theme_bw() +
    theme(text = element_text(size = 8),
          panel.grid.major = element_line(size = 0.2),
          panel.grid.minor = element_blank()))

# ggsave("Rank_abundance_curve_four_cutoffs_logy.pdf", width = 8.5, height = 8, units = "cm", p, scale = 1)


# y normal scale
(p <- ggplot(df, aes(x = Var2, y = value, group = Var1)) +
    geom_line(size = 0.4, alpha = 0.6, color = "gray") +
    geom_hline(yintercept = c(1.45), linetype = "dashed", size = 0.4, color = "#C6DBEF") + 
    geom_hline(yintercept = c(14.5), linetype = "dashed", size = 0.4, color = "#6BAED6") + 
    geom_hline(yintercept = c(1450), linetype = "dashed", size = 0.4, color = "#2171B5") + 
    geom_hline(yintercept = c(mean(cutoffs$Rarity.cutoffs)), linetype = "dashed", size = 0.4, color = "#FFCC99") + 
    annotate("text", x = 2800, y = 2, color = "#C6DBEF", label = "Rarity cutoff < 0.01 %", size = 6/.pt) + 
    annotate("text", x = 2800, y = 20, color = "#6BAED6", label = "Rarity cutoff < 0.1 %", size = 6/.pt) +
    annotate("text", x = 2800, y = 1500, color = "#2171B5", label = "Rarity cutoff < 1.0 %", size = 6/.pt) +
    annotate("text", x = 2800, y = 65, color = "#FFCC99", label = "Sample-specific rarity cutoffs ~ 0.3 %", size = 6/.pt) + 
    labs(x = "Rank", y= "Reads", title= "") +
    theme_bw() +
    theme(text = element_text(size = 8),
          panel.grid.major = element_line(size = 0.2),
          panel.grid.minor = element_blank()))

# ggsave("Rank_abundance_curve_four_cutoffs.pdf", width = 8.5, height = 8, units = "cm", p, scale = 1)

```


### Session Info
```{r}
sessionInfo()
```