---
title: "format tables for the CTS dataset"
author: "Xiu Jia (xibeihenai@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

The composted tannery sludge (CTS) dataset are from Prof. Ademir Araujo <asfaruaj@yahoo.com.br> from Federal University of Piauí. 
Sequence data was analyzed by Dr. Lucas William Mendes <lucaswmendes@gmail.com> from University of Sao Paulo.

***

##### Treatment by composted tannery sludge (CTS):
+ T1 - 0 ton CTS/ha
+ T2 - 2.5 ton CTS/ha
+ T3 - 5 ton CTS/ha
+ T4 - 10 ton CTS/ha
+ T5 - 20 ton CTS/ha

##### Sampling time points:
+ 0 - 1 days after CTS application
+ 45 - 45 days after CTS application
+ 75 - 75 days after CTS application
+ 150 - 150 days after CTS application
+ 180 - 180 days after CTS application


### initiate libraries
```{r load_packages_01, message=FALSE, warning=FALSE}

rm(list=ls())

library(tidyverse)
library(vegan)
library(picante)

```


### load OTU table
```{r}

# feature table
com <- read.csv("OTU_table_unrare.csv", header = 1, row.names = 1, sep = ",") 
dim(com)

# remove Unassigned, Archaea, Chloroplast and Mitochondria and only keep OTU assigned at phylum level

com <- com %>% 
  filter(!grepl('Archaea', taxonomy)) %>% 
  filter(!grepl('Chloroplast', taxonomy)) %>% # D_0__Bacteria; D_1__Cyanobacteria; D_2__Oxyphotobacteria; D_3__Chloroplast
  filter(!grepl('Mitochondria', taxonomy)) %>%  # D_0__Bacteria; D_1__Proteobacteria; D_2__Alphaproteobacteria; D_3__Rickettsiales; D_4__Mitochondria
  filter(grepl('D_1__', taxonomy)) #   filter(!grepl('Unassigned', taxonomy)) %>%  is not nessary

dim(com)

# write.csv(com, "out_table_bacteria_only.csv")

# extracted a taxonomy table
taxa <- com %>% 
  select(taxonomy)

dim(taxa)

# exported the taxonomy table
# write.csv(taxa, "taxonomy.csv")

# only keep OTUs from the table
com <- com %>% 
  select(-taxonomy)

dim(com)
```


### rarefy the OTU table
```{r}
# check the total sequence of samples
head(sort(colSums(com)))
dim(com)
range(rowSums(com))

# check how the table looks like
com[1:3, 1:3]

# Set cutoff by youself
cutoff <- 14500

# rarefy
set.seed(123)
rare <- rrarefy(t(com), cutoff)
rare <- rare[, colSums(rare)!=0]
rare <- t(rare)

# # after rarefying, we lost 105 features, which might have low fequency and abundance (I need to further verify this)
dim(rare)

# check if rarefy work
range(rowSums(rare))
head(sort(colSums(rare)))
rare[1:2, 1:2]

# save the rarefied table for later use
# write.csv(rare, "out_table_rarefied.csv")

```


### load phylogenetic tree
```{r eval=FALSE, warning=FALSE, include=FALSE}

phylo <- read.tree("tree.nwk")

str(phylo)

print("prune tree")
match.phylo.com <- match.phylo.data(phylo, rare); # species as rows, samples as columns for com table
str(match.phylo.com)

# save the pruned tree
# write.tree(match.phylo.com$phy, file = "tree_pruned.nwk")

```


### Session Info
```{r}
sessionInfo()
```