---
title: "Rare biosphere analysis for the CTS dataset"
author: "Xiu Jia (xibeihenai@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

The composted tannery sludge (CTS) dataset from Prof. Ademir Araujo <asfaruaj@yahoo.com.br> from Federal University of Piauí. 
Sequence data was analyzed by Dr. Lucas William Mendes <lucaswmendes@gmail.com> from University of Sao Paulo.

***

##### Treatment by composted tannery sludge (CTS) / CTS concentration:
* T1 - 0 ton CTS/ha
* T2 - 2.5 ton CTS/ha
* T3 - 5 ton CTS/ha
* T4 - 10 ton CTS/ha
* T5 - 20 ton CTS/ha

##### Sampling time points:
* 0 - 1 days after CTS application
* 45 - 45 days after CTS application
* 75 - 75 days after CTS application
* 150 - 150 days after CTS application
* 180 - 180 days after CTS application


### initiate libraries
```{r setup, message=FALSE, warning=FALSE}

rm(list=ls())

library(tidyverse)
library(vegan)
library(ape)
library(RColorBrewer)
library(ggpubr)
library(ggpmisc)

mytheme <- theme_bw()+
    theme(panel.spacing = unit(0, "lines"),
          strip.background = element_blank(),
          strip.placement = "outside",
          legend.box.background = element_rect(),
          legend.box.margin = margin(1, 1, 1, 1),
          legend.position = "right", 
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank())
```


### load OTU table
```{r}

# feature table
com <- read.csv("out_table_rarefied.csv", header = 1, row.names = 1, sep = ",") 
com <- t(com)
str(com)

knitr::kable(
  com[1:5, 1:2],
  caption = "how the OTU table looks"
)

cat("the OTU/feature table has been rarefied to", rarefactiondepth <- mean(rowSums(com)))

```
The number of samples is: `r nrow(com)`; the number of species/ASVs is: `r ncol(com)`; the range of sequence number among samples is: `r range(colSums(com))`.


### load metadata
```{r}

metadata <- read.csv("metadata.csv", header = 1, row.names = 1, sep = ",")

dim(metadata)

head(metadata)

```


### load taxonomy
```{r}

taxonomy <- read.csv("taxonomy_splited_columns.csv", header = 1, row.names = 1, sep = ",")

dim(taxonomy)

head(taxonomy)

```


### using sample specific cutoffs to define the rare and common biospheres
For more details about 'sample specific cutoffs' please check Jia, X., Dini-Andreote, F. & Salles, J.F. Unravelling the interplay of ecological processes structuring the bacterial rare biosphere. ISME COMMUN. 2, 96 (2022). https://doi.org/10.1038/s43705-022-00177-6
```{r}

# source the TruncateTable FUNCTION
source("TruncateTable.r")

# The truncated datasets can be stored as follows: 
dominant <- TruncateTable(com, typem = "dominant") 
str(dominant)

# write.csv(t(dominant), "OTU_table_dominant.csv")

rare <-TruncateTable(com, typem = "rare") 
str(rare)

# write.csv(t(rare),  "OTU_table_rare.csv")


# Combine rare and dominant biosphere
row.names(dominant) <- paste("dominant", row.names(dominant), sep = "_")
dominant <- as.data.frame(t(dominant))

row.names(rare) <- paste("rare", row.names(rare), sep = "_")
rare <- as.data.frame(t(rare))

# combine rare & dominant biospheres
com_rare_dominant <- transform(merge(dominant, rare, by = "row.names", all = TRUE), row.names = Row.names, Row.names = NULL)  
com_rare_dominant[is.na(com_rare_dominant)] <- 0
dim(com_rare_dominant)

```


### total relative abundance of the rare and dominant biosphere

```{r}
com_rare_dominant[1:2, 1:3]

df <- as.data.frame(colSums(com_rare_dominant))
df$ab <- df[, 1]
                 
df <- df %>% 
  rownames_to_column(var = "col") %>% 
  separate(col, into = c("Biosphere", "Treat", "Day"), sep = "_") %>% 
  mutate(Day = gsub('.{1}$', '', Day)) %>% 
  select(-4)

# calculate relative abundance
df$ab <- df$ab*100/rarefactiondepth

# change Treat info
df1 <- df %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Biospheres = factor(Biosphere, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere")))

df1$Day <- as.numeric(as.character(df1$Day))

head(df1)

# line plot
(p <- ggplot(df1, aes(x = Day, y = ab)) + 
    geom_smooth(method = lm, se = TRUE, size = .6, aes(color = Treat)) +
    geom_point(aes(color = Treat), size = 3, alpha = .8) + 
    scale_color_brewer(palette = "PuBu", name = "CTS concentration") +
    facet_grid(Biospheres ~ Treat, scales = "free_y") +
    stat_poly_eq(mapping = use_label(c("R2", "P")), p.digits = 2, label.y = "bottom", label.x = "left") +
    scale_x_continuous(breaks = c(0, 45, 75, 150, 180)) +
    labs(title = "", x = "Day", y = "Relative abundance (%)") +
    mytheme +
    theme(text = element_text(size = 15)))

# ggsave("Total_relative_abundance_rare_dominant_biospheres.pdf", width = 20, height = 8, units = "cm", p, scale = 1.7)
# ggsave("Total_relative_abundance_rare_dominant_biospheres.jpg", width = 20, height = 8, units = "cm", p, scale = 1.7)

# change Treat info
df2 <- df %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0", "2.5", "5", "10", "20"))) %>%
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Biospheres = factor(Biosphere, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere")))

df2$Treat <- as.numeric(as.character(df2$Treat))

head(df2)

# line plot
(p <- ggplot(df2, aes(x = Treat, y = ab)) + 
    geom_smooth(method = lm, se = TRUE, size = .6, aes(color = factor(Treat))) +
    geom_point(aes(color = factor(Treat)), size = 3, alpha = .8) + 
    scale_color_brewer(palette = "PuBu", name = "CTS concentration") +
    stat_poly_eq(mapping = use_label(c("R2", "P")), p.digits = 2, label.y = "bottom", label.x = "left") +
    scale_x_continuous(breaks = c(0, 2.5, 5, 10, 20)) +
    facet_grid(Biospheres ~ Day, scales = "free_y") +
    labs(title = "", x = "Treatment", y = "Relative abundance (%)") +
    mytheme +
    theme(text = element_text(size = 15)))

# ggsave("Total_relative_abundance_rare_dominant_biospheres_treat.pdf", width = 20, height = 8, units = "cm", p, scale = 1.7)
# ggsave("Total_relative_abundance_rare_dominant_biospheres_treat.jpg", width = 20, height = 8, units = "cm", p, scale = 1.7)

```


### number of OTUs in the rare and dominant biosphere
```{r}

richness <- specnumber(t(com_rare_dominant))
df <- data.frame(richness)

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df), t(as.data.frame(strsplit(as.character(row.names(df)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)


# combine metadata to the dominant phyla table
df <- data.frame(Biosphere = as.factor(group_info[,1]),
                 Treat = as.factor(group_info[,2]),
                 Day = as.factor(group_info[,3]),
                 df)

#  get average per group
df1 <- df %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Biospheres = factor(Biosphere, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere"))) %>%
  group_by(Biospheres, Treat, Day) %>% 
  summarise(avg = mean(richness, na.rm=T)) %>% 
  arrange(Day, rev(Biospheres)) %>% 
  group_by(Day, Treat) %>% 
  mutate(label_y = cumsum(avg) - 0.5 * avg)

df1$avg <- as.integer(df1$avg)


# stacked-bar plot
(p <- ggplot(df1, aes(x = Day, y = avg, fill = Biospheres)) + 
    geom_col(color = "black", width = 0.8, lwd = 0.1) +
    scale_y_continuous(expand = c(0, 8)) +
    facet_grid(. ~ Treat) +
    labs(x = "Days", y = "Richness (NO. of OTUs)", title = "") +
    geom_text(aes(y = label_y, label = avg),  colour = "white", size = 2) +
    scale_fill_brewer(palette = "Pastel1") +
    mytheme)

# ggsave("Richness_stacked_rare_dominant.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_1.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_1.jpg", width = 14, height = 8.5, units = "cm", p, scale = 1.5)

# line plot
df2 <- df %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Biospheres = factor(Biosphere, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere")))

pd <- position_dodge(0.2)  

(p <- ggplot(df2, aes(x = Day, y = richness, group = Treat, colour = Treat)) + 
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, alpha = 0.5, position = pd) +
    stat_summary(fun = "mean", geom = "line", alpha = 0.5, position = pd) + 
    stat_summary(fun = "mean", geom = "point", size = 3, position = pd) +
    scale_color_brewer(palette="PuBu", name = "CTS concentration") +
    facet_wrap(.~Biospheres, scales = "free_y") +
    labs(title = "", x = "Day", y = "Richness (NO. of OTUs)") +
    mytheme)

# ggsave("Richness_lineplot_freey_rare_dominant.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)

(p <- ggplot(df2, aes(x = Day, y = richness, group = Treat, colour = Treat)) + 
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, alpha = 0.5, position = pd) +
    stat_summary(fun = "mean", geom = "line", alpha = 0.5, position = pd) + 
    stat_summary(fun = "mean", geom = "point", size = 3, position = pd) +
    scale_color_brewer(palette="PuBu", name = "CTS concentration") +
    facet_grid(.~Biospheres) +
    labs(title = "", x = "Day", y = "Richness (NO. of OTUs)") +
    mytheme)

# ggsave("Richness_lineplot_rare_dominant.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_1_optional.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_1_optional.jpg", width = 14, height = 8.5, units = "cm", p, scale = 1.5)

```


### beta-diversity

```{r fig.show = "hold", out.width = "50%"}

# dominant biosphere
dist <- vegdist(t(dominant), method = "bray", binary = FALSE, diag = 1) 

re <- pcoa(dist, correction = "none", rn = NULL)
df <- re$vectors[, 1:2]

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df), t(as.data.frame(strsplit(as.character(row.names(df)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)

# combine metadata to the dominant phyla table
df <- data.frame(Biosphere = as.factor(group_info[,1]),
                 Treat = as.factor(group_info[,2]),
                 Day = as.factor(group_info[,3]),
                 df)

df <- df %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) 

# two way permanova
set.seed(123)
result <- adonis(dist ~ Treat*Day, data = df,  permutation = 9999) 
result$aov.tab

# write.csv(result$aov.tab, "PERMANOVA_dominant.csv")


(p1 <- ggplot(df, aes(Axis.1, Axis.2, shape = Treat, color = Day)) +
    geom_point(size = 3, alpha = 0.7) + 
    labs(x = paste("PCoA1 (", round(re$values$Relative_eig[1] * 100, 2), "%)", sep = ""), 
         y = paste("PCoA2 (", round(re$values$Relative_eig[2] * 100, 2), "%)", sep = ""), 
         title = "Dominant biosphere") +
    scale_color_brewer(palette="Dark2") +
    scale_shape_manual(values = c(1, 0, 15, 16, 17), name = "CTS concentration") +
    theme_bw())


# rare biosphere
dist <- vegdist(t(rare), method = "bray", binary = FALSE, diag = 1) 

re <- pcoa(dist, correction = "none", rn = NULL)
df <- re$vectors[, 1:2]

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df), t(as.data.frame(strsplit(as.character(row.names(df)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)

# combine metadata to the dominant phyla table
df <- data.frame(Biosphere = as.factor(group_info[,1]),
                 Treat = as.factor(group_info[,2]),
                 Day = as.factor(group_info[,3]),
                 df)

df <- df %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) 

# two way permanova
set.seed(123)
result <- adonis(dist ~ Treat*Day, data = df,  permutation = 9999) 
result$aov.tab

# write.csv(result$aov.tab, "PERMANOVA_rare.csv")

(p2 <- ggplot(df, aes(Axis.1, Axis.2, shape = Treat, color = Day)) +
    geom_point(size = 3, alpha = 0.7) + 
    labs(x = paste("PCoA1 (", round(re$values$Relative_eig[1] * 100, 2), "%)", sep = ""), 
         y = paste("PCoA2 (", round(re$values$Relative_eig[2] * 100, 2), "%)", sep = ""), 
         title = "Rare biosphere") +
    scale_color_brewer(palette="Dark2") +
    scale_shape_manual(values = c(1, 0, 15, 16, 17), name = "CTS concentration") +
    theme_bw())


p <- ggarrange(p1, p2, labels = c("A", "B"), common.legend = TRUE, legend = "right", ncol = 2)

# ggsave("PCoA_rare_dominant.pdf", width = 16, height = 7.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_2.pdf", width = 16, height = 7.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_2.jpg", width = 16, height = 7.5, units = "cm", p, scale = 1.5)

```


### Phyla composition
```{r}

# species composition of the rare and dominant biosphere at Phyla level (relative abundance) 

# add taxonomy info
df <- transform(merge(taxonomy[, -c(1, 4:7)], com_rare_dominant, by = "row.names"), row.names = Row.names, Row.names = NULL)  

df <- df %>% select(-Class)


warning("double check is the table rarified or not?")
(rarefactiondepth <- mean(rowSums(com)))

# combine OTUs in the same phylum
df <- aggregate(df[,-1], list(df$Phylum), sum)
row.names(df) <- df$Group.1
df <- df[, -1]
df <- 100*df/rarefactiondepth


df[1:2, 1:4]
row.names(df)

```


all phyla
```{r}
# transpose the dataframe
df1 <- t(df)
df1 <- as.data.frame(df1)

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df1), t(as.data.frame(strsplit(as.character(row.names(df1)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)

# combine metadata to the dominant phyla table
df1.1 <- data.frame(Biosphere = as.factor(group_info[,1]),
                  Treat = as.factor(group_info[,2]),
                  Day = as.factor(group_info[,3]),
                  df1)

#  pivot data frame from wide to long
df1.2 <- df1.1 %>%
  pivot_longer(cols = 4:ncol(df1.1), 
               names_to = "Phyla", 
               values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Biosphere = factor(Biosphere, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere")))

head(df1.2)

# write_csv(df1.2, "phyla_composition_all_rare_dominnat.csv")

#  pivot data frame from wide to long
df1.1 <- df1.1 %>%
  pivot_longer(cols = 4:ncol(df1.1), 
               names_to = "Phyla", 
               values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Biosphere = factor(Biosphere, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere"))) %>% 
  group_by(Biosphere, Treat, Day, Phyla) %>% 
  summarise(avg = mean(value, na.rm=T)) 

head(df1.1)

set.seed(112358)
(colourCount = length(colnames(df1)))
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col <- sample(col_vector, colourCount)
# pie(rep(1,colourCount), col=sample(col_vector, colourCount))

# stacked-bar plot
(p <- ggplot(df1.1, aes(x = Day, y = avg, fill = Phyla)) + 
    geom_bar(stat = "identity", width = 0.8) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values = col) +
    facet_grid(Biosphere ~ Treat) +
    labs(x = "Days", y = "Relative abundance (%)", title = "") +
    guides(fill = guide_legend(title="Phyla", reverse = TRUE)) +
    mytheme)

ggsave("phyla_stacked_barplots_all_rare_dominant.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_3.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_3.jpg", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
```


only keep phyla with average relative abundance > 0.1%
```{r}
# only keep phyla with average relative abundance > 0.1%
df2 <- df
df2$mean <- rowMeans(df2)
df2 <- subset(df2, mean > 0.1)
df2 <- df2[with(df2, order(mean)), ]
df2$mean <- NULL

# phyla that abundance more than 0.1%
row.names(df2)

# transpose the dataframe
df2 <- t(df2)
df2 <- as.data.frame(df2)

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df2), t(as.data.frame(strsplit(as.character(row.names(df2)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)

# combine metadata to the dominant phyla table
df3 <- data.frame(Biosphere = as.factor(group_info[,1]),
                  Treat = as.factor(group_info[,2]),
                  Day = as.factor(group_info[,3]),
                  df2)

#  pivot data frame from wide to long
df3 <- df3 %>%
  pivot_longer(cols = 4:ncol(df3), 
               names_to = "Phyla", 
               values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Biosphere = factor(Biosphere, levels = c("dominant", "rare"), labels = c("Dominant biosphere", "Rare biosphere"))) %>% 
  group_by(Biosphere, Treat, Day, Phyla) %>% 
  summarise(avg = mean(value, na.rm=T)) #  summarise_at(vars("PM25", "Ozone", "CO2"), mean); or summarise(across(PM25:CO2, mean))

head(df3)

set.seed(112358)
(colourCount = length(colnames(df2)))
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col <- sample(col_vector, colourCount)
# pie(rep(1,colourCount), col=sample(col_vector, colourCount))

# stacked-bar plot
(p <- ggplot(df3, aes(x = Day, y = avg, fill = Phyla)) + 
    geom_bar(stat = "identity", width = 0.8) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values = col) +
    facet_grid(Biosphere ~ Treat) +
    labs(x = "Days", y = "Relative abundance (%)", title = "") +
    guides(fill = guide_legend(title="Phyla", reverse = TRUE)) +
    mytheme)

# ggsave("phyla_stacked_barplots_more_than_0.1_rare_dominant.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_3.pdf", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_3.jpg", width = 14, height = 8.5, units = "cm", p, scale = 1.5)
```

### define different types of rarity and commonness

first find the most types (except transiently rare)
```{r}
# find all taxa belongs to the rare biosphere
taxa_rare <- row.names(rare)
taxa_dominant <- row.names(dominant)

# create a data.frame that
df_types <- data.frame(OTU_ID = colnames(com), 
                 Types = NA)

conditionally_rare_dominant <- taxa_dominant[taxa_dominant %in% taxa_rare]
permanently_dominant <- taxa_dominant[!taxa_dominant %in% taxa_rare]
only_rare <- taxa_rare[!taxa_rare %in% taxa_dominant]

cat("the number of OTUs in the dominant biosphere is", length(taxa_dominant),
    "\nthe number of OTUs in the rare biosphere is", length(taxa_rare),
    "\nthe number of conditionally rare/dominant OTUs is", length(conditionally_rare_dominant),
    "\nthe number of OTUs that only apears in the dominant biosphere is", length(permanently_dominant),
    "\nthe number of rare OTUs that not appear in the dominant biosphere is", length(only_rare))


df_types$Types[df_types$OTU_ID %in% conditionally_rare_dominant] <- "conditionally_rare_dominant"
df_types$Types[df_types$OTU_ID %in% only_rare] <- "only_rare"
df_types$Types[df_types$OTU_ID %in% permanently_dominant] <- "permanently_dominant"

df_types <- df_types %>%
  # mutate(Types = factor(Types)) %>% 
  mutate(OTU_ID = as.character(OTU_ID))

head(df_types)
table(df_types$Types)

# write.csv(df_types,  "OTU_ID_and_types.csv")
```


different types of rarity and dominance in a facet plot
```{r}

com_classified <- com_rare_dominant %>% 
  rownames_to_column(var = "OTU_ID") 

com_classified <- left_join(df_types, com_classified, by = "OTU_ID") %>% 
  column_to_rownames(var = "OTU_ID")


# last step - define the "transiently rare"
com_classified$Types[which(rowSums(com_classified[, -1] != 0) == 1 & com_classified$Types != "conditionally_rare_dominant "& com_classified$Types != "permanently_dominant") ] <- "transiently_rare"

com_classified[1:2, 1:3]

df <- com_classified %>%
  group_by(Types) %>%
  summarise(across(everything(), list(sum))) %>% 
  column_to_rownames(var = "Types")

df <- as.data.frame(t(df))

# calculate relative abundance
df <- 100*df/rarefactiondepth

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df), t(as.data.frame(strsplit(as.character(row.names(df)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)

# combine metadata to the dominant phyla table
df1 <- data.frame(Biosphere = as.factor(group_info[,1]),
                  Treat = as.factor(group_info[,2]),
                  Day = as.factor(group_info[,3]),
                  df)

#  pivot data frame from wide to long
df2 <- df1 %>%
  pivot_longer(cols = 4:ncol(df1), 
               names_to = "Types", 
               values_to = "value") 

df2$Types[df2$Types == "conditionally_rare_dominant" & df2$Biosphere == "dominant"] <- "conditionally_dominant"
df2$Types[df2$Types == "conditionally_rare_dominant" & df2$Biosphere == "rare"] <- "conditionally_rare"

df2 <- df2 %>% 
  select(-Biosphere) %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Types = factor(Types, levels = c("conditionally_dominant", "permanently_dominant", "only_rare", "conditionally_rare", "transiently_rare"), labels = c("Conditionally dominant", "Permanently dominant", "Permanently rare", "Conditionally rare", "Transiently rare"))) %>% 
  filter(value > 0)

df2$Day <- as.numeric(as.character(df2$Day))

head(df2)
str(df2)

# line plot
(p <- ggplot(df2, aes(x = Day, y = value)) + 
    geom_smooth(method = lm, se = TRUE, size = .6, aes(color = Treat)) +
    geom_point(aes(color = Treat), size = 3, alpha = .8) + 
    scale_color_brewer(palette = "PuBu", name = "CTS concentration") +
    facet_grid(Types~Treat, scales="free_y") +
    stat_poly_eq(mapping = use_label(c("R2", "P")), p.digits = 2, label.y = "bottom", label.x = "left") +
    scale_x_continuous(breaks=c(0, 45, 75, 150, 180)) +
    labs(title = "", x = "Day", y = "Relative abundance (%)") +
    mytheme +
    theme(text = element_text(size = 15)))

# ggsave("Figure_5_new.pdf", width = 20, height = 15, units = "cm", p, scale = 1.7)
# ggsave("Figure_5_new.jpg", width = 20, height = 15, units = "cm", p, scale = 1.7)


```


dominant biosphere
```{r}

# add rarify classification to the dominant OTU table
dominant_classified <- dominant %>% 
  rownames_to_column(var = "OTU_ID") 

dominant_classified <- left_join(dominant_classified, df_types, by = "OTU_ID") %>% 
  column_to_rownames(var = "OTU_ID")

# define the "transiently dominant"
# dominant_classified$Types[which(rowSums(dominant_classified[,-ncol(dominant_classified)] != 0) == 1)] = "transiently_dominant"
# conditionally_rare_dominant 
#                         236 
#        permanently_dominant 
#                           1 
#        transiently_dominant 
#                         234 

# write.csv(dominant_classified,  "OTU_table_dominant_with_types.csv")

# summarize sequence number per type
df <- aggregate(dominant_classified[,-ncol(dominant_classified)], list(dominant_classified$Types), sum)

row.names(df) <- df$Group.1
df <- df[, -1]
df <- as.data.frame(t(df))

# calculate relative abundance
df <- 100*df/rarefactiondepth

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df), t(as.data.frame(strsplit(as.character(row.names(df)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)

# combine metadata to the dominant phyla table
df1 <- data.frame(Biosphere = as.factor(group_info[,1]),
                  Treat = as.factor(group_info[,2]),
                  Day = as.factor(group_info[,3]),
                  df)

#  pivot data frame from wide to long
df2 <- df1 %>%
  pivot_longer(cols = 4:ncol(df1), 
               names_to = "Types", 
               values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Types = factor(Types, levels = c("conditionally_rare_dominant", "permanently_dominant"), labels = c("Conditionally dominant", "Permanently dominant"))) %>% 
  group_by(Biosphere, Treat, Day, Types) %>% 
  summarise(avg = mean(value, na.rm=T)) 

df2$avg <- as.integer(df2$avg)

my_palette = c(brewer.pal(8, "Pastel2")[c(5:6)])

# stacked-bar plot
(p1 <- ggplot(df2, aes(x = Day, y = avg, fill = Types)) + 
    geom_col(color = "black", width = 0.8, lwd = 0.1) +
    scale_y_continuous(expand = c(0, 0), limits=c(0, 80)) +
    scale_fill_manual(values = my_palette) +
    facet_grid(. ~ Treat) +
    labs(x = "Days", y = "Relative abundance (%)", title = "Dominant biosphere") +
    guides(fill = guide_legend(title="Types of dominance")) +
    mytheme+
    theme(legend.position = "bottom"))

# line plot

#  pivot data frame from wide to long
df3 <- df1 %>%
  pivot_longer(cols = 4:ncol(df1), 
               names_to = "Types", 
               values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Types = factor(Types, levels = c("conditionally_rare_dominant", "permanently_dominant"), labels = c("Conditionally dominant", "Permanently dominant")))

(f1 <- ggplot(df3, aes(x = Day, y = value, group = Treat, colour = Treat)) + 
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, alpha = 0.5, position = pd) +
    stat_summary(fun = "mean", geom = "line", alpha = 0.4, position = pd) + 
    stat_summary(fun = "mean", geom = "point", size = 2, alpha = 0.9, position = pd) +
    scale_color_brewer(palette="PuBu", name = "CTS concentration") +
    facet_grid(.~Types) +
    labs(title = "Dominant biosphere", x = "Day", y = "Relative abundance (%)") +
    mytheme)

```

rare biosphere
```{r}

# add rarify classification to the rare OTU table
rare_classified <- rare %>% 
  rownames_to_column(var = "OTU_ID") 

rare_classified <- left_join(rare_classified, df_types, by = "OTU_ID") %>% 
  column_to_rownames(var = "OTU_ID")

# define the "transiently rare"
rare_classified$Types[which(rowSums(rare_classified[,-ncol(rare_classified)] != 0) == 1)] = "transiently_rare"

# write.csv(rare_classified,  "OTU_table_rare_with_types.csv")

# summarize sequence number per type
df <- aggregate(rare_classified[,-ncol(rare_classified)], list(rare_classified$Types), sum)

row.names(df) <- df$Group.1
df <- df[, -1]
df <- as.data.frame(t(df))

# calculate relative abundance
df <- 100*df/rarefactiondepth

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df), t(as.data.frame(strsplit(as.character(row.names(df)), "_"))))

# remove last letter in the third column
group_info$X3 <- gsub('.{1}$', '', group_info$X3)

# combine metadata to the dominant phyla table
df1 <- data.frame(Biosphere = as.factor(group_info[,1]),
                  Treat = as.factor(group_info[,2]),
                  Day = as.factor(group_info[,3]),
                  df)

#  pivot data frame from wide to long
df2 <- df1 %>%
  pivot_longer(cols = 4:ncol(df1), 
               names_to = "Types", 
               values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Types = factor(Types, levels = c("only_rare", "transiently_rare", "conditionally_rare_dominant"), labels = c("Permanently rare", "Transiently rare", "Conditionally rare"))) %>% 
  group_by(Biosphere, Treat, Day, Types) %>% 
  summarise(avg = mean(value, na.rm=T)) 

df2$avg <- as.integer(df2$avg)

my_palette = c(brewer.pal(8, "Pastel1")[c(1:3)])

# stacked-bar plot
(p2 <- ggplot(df2, aes(x = Day, y = avg, fill = Types)) + 
    geom_col(color = "black", width = 0.8, lwd = 0.1) +
    scale_y_continuous(expand = c(0, 0), limits=c(0, 80)) +
    scale_fill_manual(values = my_palette) +
    facet_grid(. ~ Treat) +
    labs(x = "Days", y = "Relative abundance (%)", title = "Rare biosphere") +
    guides(fill = guide_legend(title="Types of rarity")) +
    mytheme +
    theme(legend.position = "bottom"))

df3 <- df1 %>%
  pivot_longer(cols = 4:ncol(df1), 
               names_to = "Types", 
               values_to = "value") %>% 
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) %>% 
  mutate(Types = factor(Types, levels = c("only_rare", "conditionally_rare_dominant", "transiently_rare"), labels = c("Permanently rare", "Conditionally rare", "Transiently rare")))

(f2 <- ggplot(df3, aes(x = Day, y = value, group = Treat, colour = Treat)) + 
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, alpha = 0.5, position = pd) +
    stat_summary(fun = "mean", geom = "line", alpha = 0.4, position = pd) + 
    stat_summary(fun = "mean", geom = "point", size = 2, alpha = 0.9, position = pd) +
    scale_color_brewer(palette="PuBu", name = "CTS concentration") +
    facet_grid(.~Types) +
    labs(title = "Rare biosphere", x = "Day", y = "Relative abundance (%)") +
    mytheme)

```


```{r}
p <- ggarrange(p1, p2, labels = c("A", "B"), ncol = 2)

# ggsave("Types_of_rare_dominant_abundance_stacked_barplots.pdf", width = 24, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_5.pdf", width = 24, height = 8.5, units = "cm", p, scale = 1.5)
# ggsave("Figure_5.jpg", width = 24, height = 8.5, units = "cm", p, scale = 1.5)

f <- ggarrange(f1, f2, labels = c("A", "B"), widths = c(0.8, 1.2), common.legend = TRUE, legend = "right", ncol = 2)

# ggsave("Types_of_rare_dominant_abundance_lineplots_raw.pdf", width = 20, height = 5, units = "cm", f, scale = 1.5)
# ggsave("Figure_6.pdf", width = 20, height = 5, units = "cm", f, scale = 1.5)
# ggsave("Figure_6.jpg", width = 20, height = 5, units = "cm", f, scale = 1.5)

```


### the abundance and taxonomy composition of conditionally rare/dominant
```{r}

df <- as.data.frame(t(com))

df <- df %>% 
  rownames_to_column(var = "OTU_ID") 

df <- left_join(df_types, df, by = "OTU_ID") %>% 
  column_to_rownames(var = "OTU_ID") %>% 
  filter(Types == "conditionally_rare_dominant") %>% 
  select(-Types)

```
There are `r nrow(df)` OTUs are conditionally rare/dominant.


Check the relative abundance of these conditionally rare/dominant at genus level
```{r}
# add taxonomy info - Genera
df <- transform(merge(taxonomy[, -c(1:5)], df, by = "row.names"), row.names = Row.names, Row.names = NULL)  

df <- df %>% select(-Species)

df <- aggregate(df[,-1], list(df$Genus), sum)
row.names(df) <- df$Group.1
df <- df[, -1]
df <- 100*df/rarefactiondepth

# remove these rows
row.names.remove <- c("", "uncultured", "uncultured soil bacterium")

df <- df[!(row.names(df) %in% row.names.remove), ]

# genus kept for plotting
row.names(df)

df <- as.data.frame(t(df))

# prepare a meta data file
group_info <- data.frame(row.names = rownames(df), t(as.data.frame(strsplit(as.character(row.names(df)), "_"))))

# remove last letter in the third column
group_info$X2 <- gsub('.{1}$', '', group_info$X2)

# combine metadata to the dominant phyla table
df <- data.frame(Treat = as.factor(group_info[,1]),
                 Day = as.factor(group_info[,2]),
                 df)

# write.csv(df, "conditionally_rare_dominant_genera.csv")

#  pivot data frame from wide to long
df1 <- df %>%
  pivot_longer(cols = 3:ncol(df), 
               names_to = "Genus", 
               values_to = "value") 

head(df1)

df1 <- df1 %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0 ton CTS/ha", "2.5 ton CTS/ha", "5 ton CTS/ha", "10 ton CTS/ha", "20 ton CTS/ha"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) 
# %>% 
#   group_by(Treat, Day, Genus) %>% 
#   summarise(avg = mean(value, na.rm=T)) 

pd <- position_dodge(0.2)  

(p <- ggplot(df1, aes(x = Day, y = value, group = Treat, colour = Treat)) + 
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, alpha = 0.5, position = pd) +
    stat_summary(fun = "mean", geom = "line", alpha = 0.5, position = pd) + 
    stat_summary(fun = "mean", geom = "point", size = 3, position = pd) +
    scale_color_brewer(palette="PuBu", name = "CTS concentration") +
    facet_wrap(~Genus, nrow = 11, scales = "free_y") +
    labs(title = "Conditionally rare/dominant genera", x = "Day", y = "Relative abundance (%)") +
    mytheme + 
    theme(legend.position = c(0.8, 0.03)) + 
    theme(strip.text = element_text(face = "italic")))

# ggsave("genus_conditionally_rare_dominant.pdf", width = 32, height = 30, units = "cm", p, scale = 1.5)

df2 <- df1 %>% filter(
  Genus %in% c("Nocardioides", "Nordella", "Parviterribacter", "Rubrobacter", "Tellurimicrobium",
  "uncultured.actinobacterium", "uncultured.Gaiella.sp.", "uncultured.Solirubrobacter.sp."))

df2

(p2 <- ggplot(df2, aes(x = Day, y = value, group = Treat, colour = Treat)) + 
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, alpha = 0.5, position = pd) +
    stat_summary(fun = "mean", geom = "line", alpha = 0.5, position = pd) + 
    stat_summary(fun = "mean", geom = "point", size = 3, position = pd) +
    scale_color_brewer(palette="PuBu", name = "CTS concentration") +
    facet_wrap(~Genus, nrow = 4, scales = "free_y") +
    labs(title = "", x = "Day", y = "Relative abundance (%)") +
    mytheme + 
    theme(strip.text = element_text(face = "italic")))

# ggsave("Figure_4.pdf", width = 10, height = 12, units = "cm", p2, scale = 1.5)
# ggsave("Figure_4.jpg", width = 10, height = 12, units = "cm", p2, scale = 1.5)


# another way to show the results

#  pivot data frame from wide to long
df1 <- df %>%
  pivot_longer(cols = 3:ncol(df), 
               names_to = "Genus", 
               values_to = "value") 

head(df1)

df1 <- df1 %>%
  mutate(Treat = factor(Treat, levels = c("T1", "T2", "T3", "T4", "T5"), labels = c("0", "2.5", "5", "10", "20"))) %>% 
  mutate(Day = factor(Day, levels = c("0", "45", "75", "150", "180"))) 

df2 <- df1 %>% filter(
  Genus %in% c("Nocardioides", "Nordella", "Parviterribacter", "Rubrobacter", "Tellurimicrobium",
  "uncultured.actinobacterium", "uncultured.Gaiella.sp.", "uncultured.Solirubrobacter.sp."))

df2
# %>% 
#   group_by(Treat, Day, Genus) %>% 
#   summarise(avg = mean(value, na.rm=T)) 

(p3 <- ggplot(df2, aes(x = Treat, y = value, group = Treat, colour = Treat)) + 
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, alpha = 0.5, position = pd) +
    stat_summary(fun = "mean", geom = "line", alpha = 0.5, position = pd) + 
    stat_summary(fun = "mean", geom = "point", size = 3, alpha = 0.8, position = pd) +
    scale_color_brewer(palette="PuBu") +
    guides(col = FALSE) +
    facet_wrap(~Genus, nrow = 4, scales = "free_y") +
    labs(title = "", x = "CTS concentration (ton CTS/ha)", y = "Relative abundance (%)") +
    mytheme + 
    theme(strip.text = element_text(face = "italic")))

# ggsave("Figure_4_new.pdf", width = 8, height = 12, units = "cm", p3, scale = 1.5)
# ggsave("Figure_4_new.jpg", width = 8, height = 12, units = "cm", p3, scale = 1.5)

```


### Session Info
```{r}
sessionInfo()
```