---
title: "Stacked bar plot at genus level -- Yu, Yi"
author: "Jia, Xiu (xibeihenai@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

This is the script for generating Figure S6 in Yu et al. Molecular Cell (2022).

### Initiate libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(reshape2)
```

### Import dataset

```{r, message=FALSE, warning=FALSE}

rm(list=ls())

# load feature/OTU table
com <- read.csv("feature-table-rarified-nontax.tsv", sep="\t", header=1, row.names=1, check.names = FALSE)

# replace hypen by underscore
colnames(com) <- gsub(colnames(com), pattern = '\\-', replacement = '_')
com <- t(com)

# only keep samples that belongs to L3 and heterozygous
com<- com[grepl("L3", row.names(com)),]
com<- com[grepl("heterozygous", row.names(com)),]
com <- t(com)
com[1:5, 1:2]

# load taxonomy table
taxa <- read.csv("taxonomy.csv", sep = ",", row.names = 1, header = TRUE)
taxa <- taxa[row.names(taxa) %in% row.names(com), ]
head(taxa)

```
In the end, we got `r ncol(com)` samples and `r nrow(com)` ASVs for the L3 stage and heterozygous samples. The total abundance per sample range: `r colSums(com)`.


### Summarize the relative abundance of each genus

Calculate relative abundance for each genus. Only keep taxa with relative abundance > 0.05% for final plotting
```{r, message=FALSE, warning=FALSE}

# combine otu table and taxa info
df <- transform(merge(taxa, com, by="row.names"), row.names=Row.names, Row.names=NULL)

# only keep Genus column
df <- df[, -c(1:5, 7, 8)]
df[1:5, 1:2]

warning("double check is the table rarified or not?")
(rarefactiondepth <- mean(colSums(com)))

# combine ASVs have same taxonomy
df <- aggregate(df[,-1], list(df$Genus), sum)
row.names(df) <- df$Group.1
df <- df[, -1]
df <- 100*df/rarefactiondepth

# only keep taxa with relative abundance > 0.05%
df$mean <- rowMeans(df)
df <- subset(df, mean > 0.05) 
df <- df[with(df, order(mean)), ]
df$mean <- NULL

```
In the end, we got the following Genus with avarage relative abundance > 0.05%: `r row.names(df)`

###  Clean the data set and prepare data for plotting

```{r}
# remove rows that was not assigend at genus level
df <- df[-9, ]
df <- t(df)
df <- as.data.frame(df)

# remove "uncultured bacterium"
df$`uncultured bacterium` <- NULL

# add a column for Others (such as ASVs that are not assigned at Genus level or Genera with abundance less than 0.1%)
df1 <- df %>% rownames_to_column('sample') 

df1 <- df1 %>%  mutate(Others = (100- rowSums(df1[, -1]))) %>%
  column_to_rownames('sample')

# move the last column to the start
df1 <- df1 %>%
  select(Others, everything())

head(df1)

```

### Torture the data to show less Acetobactor

As Acetobactor is the dominant genus, the variation of other genera is hard to see. So we deduct a number (60%) for the relative abundance of Acetobactor for plotting. Later on, we add a break in the y-axis manualy in Adobe Illustrator.
```{r}
df1$Acetobacter <- df1$Acetobacter - 60

# melt data.frame
df1 <- melt(as.matrix(df1))

# add group information
df1 <- df1 %>%
  separate(Var1, c("Stage", "Treatments", "Genotype", "Replicates"), "_")

df1$Treatments[df1$Treatments == 'VitaminB5'] <- 'Pantothenate'

df1$Groups <-  factor(paste(df1$Treatments, df1$Replicates, sep = "_"), level = c("Pantothenate_1", "Pantothenate_2", "Pantothenate_3", "Pantothenate_4", "Pantothenate_5", "Pantothenate_6", "Pantothenate_7", "Pantethine_1", "Pantethine_2",  "Pantethine_3",  "Pantethine_4","Pantethine_5",  "Pantethine_6", "Pantethine_7","Pantethine_8", "Control_1", "Control_2", "Control_3",    "Control_4", "Control_5", "Control_6", "Control_7", "Control_8"))

head(df1)
str(df1)
```


### Stacked bar plot 

```{r}

# colour code
set.seed(2333)
(colourCount = length(unique(df1$Var2)))
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col <- sample(col_vector, colourCount)
# pie(rep(1,colourCount), col=sample(col_vector, colourCount))

# stacked-bar plot
(p <- ggplot(df1, aes(x = Groups, y = value, fill = Var2)) + 
    geom_bar(stat = "identity", width = 0.8) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values = col) +
    labs(x = "", y = "Relative abundance (%)", title = "L3 stage (heterozygous)") +
    guides(fill = guide_legend(title="Genera", reverse = TRUE)) +
    coord_flip() +
    theme_bw()+
    theme(text = element_text(size = 10),
          panel.spacing = unit(0, "lines"),
          strip.background = element_blank(),
          strip.placement = "outside",
          legend.box.background = element_rect(),
          legend.box.margin = margin(1, 1, 1, 1),
          legend.text = element_text(face = "italic"),
          legend.position = "right", 
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.margin=unit(c(1, 0.5, 1, 0.5),"cm")))

ggsave("stacked_barplots_L3_heterozygous_genera_more_than_0.05.pdf", width = 12, height = 8.5, units = "cm", p, scale = 1.5) 

```

@Yi Please edit the raw figure in AI by adding the break in the x-axis. Where relative abundance is 10%, it should be 70%.

### Session Info

```{r}
sessionInfo()
```
